Think deep for reef resilience

### 2025-06-26 paper/paper_draft.qmd
Add carbonate budget analysis plots and clean up notes on scripts

map_width = fig_sizes["map_width"]
ecs_width = fig_sizes["ecs_width"]
timeseries_width = fig_sizes["timeseries_width"]
violin_width = fig_sizes["violin_width"]
```

@@

ADRIA output absolute coral cover was used to calculate relative coral cover (relative to initial cover), a reef’s absolute coral cover at each timestep divided by initial coral cover. Initial coral cover is consistent throughout all simulated and GCM scenarios. To limit the influence of reefs with very low initial cover on the results, reefs with future total cover levels that exceed five times their initial cover level were removed from clustering and further analyses. The proportion of each functional group occupying the reef (taxa evenness) was also investigated as a metric for timeseries analysis, but was found to not sufficiently capture the declining condition of reefs suffering from very low absolute coral cover.

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef relative coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries [@batista2014]. These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [@steinmann2020].

@@

@@

Conversely, there are no consistent patterns across bioregions in reef relative cover and mean timeseries DHW levels, with clusters having similar mean DHW levels in many bioregions (@fig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean; @sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean; @sfig-noresm2-mm-bioregion-dhwmean).

@@

Within the four management areas, timeseries clustering analysis identified three clusters of reefs using relative coral cover timeseries. These clusters form groups of reefs with ‘low, medium and high’ relative coral cover, categorised by their median coral cover between 2030 and 2040. The ‘high’ cover cluster of reefs contains reefs that have continuing recovery of corals throughout the future climate scenario. The medium and low clusters contain many reefs that decrease to half their initial coral cover levels and largely do not display the levels of recurring recovery that reefs in high clusters experience (@fig-access-esm1-5-manarea-timeseries; @sfig-access-cm2-manarea-timeseries; @sfig-ec-earth3-veg-manarea-timeseries; @sfig-gfdl-cm4-manarea-timeseries; @sfig-noresm2-mm-manarea-timeseries).

Within each bioregion reef median depth level increases with increasing reef relative coral cover. Reefs in the low cluster have the lowest median depth levels, while reefs in the high cluster have the highest median depth levels within each management area (@fig-access-esm1-5-manarea-depth; @sfig-access-cm2-manarea-depth; @sfig-ec-earth3-veg-manarea-depth; @sfig-gfdl-cm4-manarea-depth; @sfig-noresm2-mm-manarea-depth).

@@

Timeseries clustering identified three clusters in reef relative cover timeseries for the whole GBR scale. A ‘low’ cluster of reefs with little recovery following initial declines, a medium cluster of reefs with median relative cover below 0.5 post 2028, and a high cluster of reefs with median relative cover above 0.5 until 2050 (@fig-access-esm1-5-gbr-timeseries; @sfig-access-cm2-gbr-timeseries; @sfig-ec-earth3-veg-gbr-timeseries; @sfig-gfdl-cm4-gbr-timeseries; @sfig-noresm2-mm-gbr-timeseries).

The clusters identified with timeseries clustering at the GBR scale show similar patterns to clusters identified at bioregion and management area scales. Reef depth (@fig-access-esm1-5-gbr-depth; @sfig-access-cm2-gbr-depth; @sfig-ec-earth3-veg-gbr-depth; @sfig-gfdl-cm4-gbr-depth; @sfig-noresm2-mm-gbr-depth) and total cover – DHW correlation (@sfig-access-esm1-5-gbr-dhwcor; @sfig-access-cm2-gbr-dhwcor; @sfig-ec-earth3-veg-gbr-dhwcor; @sfig-gfdl-cm4-gbr-dhwcor; @sfig-noresm2-mm-gbr-dhwcor) decrease, while the proportion of coral-habitable area initially occupied by corals increases (@sfig-access-esm1-5-gbr-initialproportion; @sfig-access-cm2-gbr-initialproportion; @sfig-ec-earth3-veg-gbr-initialproportion; @sfig-gfdl-cm4-gbr-initialproportion; @sfig-noresm2-mm-gbr-initialproportion), as the clusters decrease from ‘high’ to ‘low’ cover levels. There are no substantial differences between the clusters in mean DHW (@sfig-access-esm1-5-gbr-dhwmean; @sfig-access-cm2-gbr-dhwmean; @sfig-ec-earth3-veg-gbr-dhwmean; @sfig-gfdl-cm4-gbr-dhwmean; @sfig-noresm2-mm-gbr-dhwmean) and DHW tolerance (@sfig-access-esm1-5-gbr-dhwtol; @sfig-access-cm2-gbr-dhwtol; @sfig-ec-earth3-veg-gbr-dhwtol; @sfig-gfdl-cm4-gbr-dhwtol; @sfig-noresm2-mm-gbr-dhwtol) levels of reefs. Similarly, there are no substantial differences between clusters in connectivity metrics source – to – sink ratio and total connectivity strength (@sfig-access-esm1-5-gbr-logsosi; @sfig-access-cm2-gbr-logsosi; @sfig-ec-earth3-veg-gbr-logsosi; @sfig-gfdl-cm4-gbr-logsosi; @sfig-noresm2-mm-gbr-logsosi; @sfig-access-esm1-5-gbr-totalconnectivity; @sfig-access-cm2-gbr-totalconnectivity; @sfig-ec-earth3-veg-gbr-totalconnectivity; @sfig-gfdl-cm4-gbr-totalconnectivity; @sfig-noresm2-mm-gbr-totalconnectivity).

@@

Timeseries analysis has been used to identify three groups of reefs with similar declining (low), recovering (high), and intermediate (medium) coral cover trajectories throughout future scenarios. Throughout all analysis scales, reefs in high cover clusters exhibit spans of recovery up to 2050, while low and medium clusters commonly remain at levels below half their initial cover from 2030. Furthermore, at all analysis scales reef clusters show similar patterns in their reef characteristics such as depth and proportion of habitable area initially occupied.

A key factor that influences the relative coral cover values on a reef is the initial total coral cover used to calculate the metric. This is seen where low and medium reefs have higher proportions of their coral-habitable area occupied by corals initially, while some high-cluster reefs exhibit absolute coral cover levels greater than their initial coral cover, leading to future relative cover values greater than one.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced, and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to calculate the degree of protection provided to a reef based on the median depth. Our results suggest that within shallow reef ecosystems (depth < 30m), depth contributes greatly to reef resilience and the ability of the reef to recover following heat stress. While similar findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses have often focussed on the potential of mesophotic reefs (> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific. The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress.

@@

Current reef restoration and intervention-investigation efforts on the Great Barrier Reef, such as deployment of coral seeding and fragment devices target reefs with suitable restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One of the main depth constraints is being able to visually identify the target site from the surface for deployment and ongoing monitoring, in combination with factors such as deployment method and tourism return value. Our results suggest that of the reefs analysed within the 1-10m depth range (1800) 87% (1569) of reefs fall into the low and medium clusters (at the Great Barrier Reef scale), with many suffering large future declines in reef health under climate change and possibly limiting the potential of interventions to deliver recovery in these locations. Reef intervention planning is a complex and challenging decision making process [@anthony2020], and our results suggest that deployment site depth is likely an important factor to consider when assessing the long-term resilience of reef sites.

@@

@@

Further analyses could build on the findings of this study by analysing the impacts of prioritising reef depth in reef restoration intervention decision making. The ADRIA platform is developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and ecosystem scenario modelling into reef restoration decision analysis. The platform allows for the investigation of different reef characteristics to be targetted in decision making and models the evolving outcomes as a result of reef interventions at various locations. The capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of waves, cyclones, CoTS and currents into reef resilience investigations [@lin2023; @mcclanahan2005; @grimaldi2022; @liu2024].

@@

>>>>

map_width = fig_sizes["map_width"]
ecs_width = fig_sizes["ecs_width"]
timeseries_width = fig_sizes["timeseries_width"]
violin_width = fig_sizes["violin_width"]
carb_width = fig_sizes["carb_width"]
```

@@

ADRIA-CoralBlox was run for 4096 *sampled* scenarios per GCM. A median absolute coral cover timeseries for each reef was calculated for each GCM. For use in timeseries clustering reef cover timeseries were scaled by the total area of the reef to represent timeseries as a percentage of total reef area (range 0 - 100%). The proportion of each functional group occupying the reef (taxa evenness) was also investigated as a metric for timeseries analysis, but was found to not sufficiently capture the declining condition of reefs suffering from very low absolute coral cover.

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries [@batista2014]. These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [@steinmann2020].

## Carbonate budget analysis

To maintain coral cover over time coral reefs require a threshold of coral cover known as a carbonate budget. Estimates of the carbonate budget required to maintain a healthy coral reef range from 10 to 20% of a reef's total area. Using median absolute cover timeseries for each reef we calculated the number of years each reef maintains coral cover above a carbonate budget threshold level. Threshold values from 10 to 20% (inclusive) were used in analyses.

@@

## Reef carbonate budget
The number of years reefs maintain coral cover above a required carbonate budget level substantially decreases as the required carbonate budget threshold (as a percentage of reef area) increases. Reefs with greater median depth levels maintain coral cover above required carbonate budget levels for longer than reefs at shallow depths [@fig-access-esm1-5-depth-carbonate]. There are no clear relationships between reef total connectivity levels and the number of years reefs maintain coral cover above carbonate budget threshold levels [@sfig-access-esm1-5-conn-carbonate].

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Point colour represents the median reef depth level in metres. Displays results from median timeseries using ACCESS-ESM1-5.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/depth_carbonate_scatter.png){#fig-access-esm1-5-depth-carbonate width=`{julia} carb_width`}

@@

Conversely, there are no consistent patterns across bioregions in reef relative cover and mean timeseries DHW levels, with clusters having similar mean DHW levels in many bioregions (@sfig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean; @sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean; @sfig-noresm2-mm-bioregion-dhwmean).

@@

Within the four management areas, timeseries clustering analysis identified three clusters of reefs using relative coral cover timeseries. These clusters form groups of reefs with ‘low, medium and high’ relative coral cover, categorised by their median coral cover between 2030 and 2040. The ‘high’ cover cluster of reefs contains reefs that have continuing recovery of corals throughout the future climate scenario. The medium and low clusters contain many reefs that decrease to half their initial coral cover levels and largely do not display the levels of recurring recovery that reefs in high clusters experience (@sfig-access-esm1-5-manarea-timeseries; @sfig-access-cm2-manarea-timeseries; @sfig-ec-earth3-veg-manarea-timeseries; @sfig-gfdl-cm4-manarea-timeseries; @sfig-noresm2-mm-manarea-timeseries).

Within each bioregion reef median depth level increases with increasing reef relative coral cover. Reefs in the low cluster have the lowest median depth levels, while reefs in the high cluster have the highest median depth levels within each management area (@sfig-access-esm1-5-manarea-depth; @sfig-access-cm2-manarea-depth; @sfig-ec-earth3-veg-manarea-depth; @sfig-gfdl-cm4-manarea-depth; @sfig-noresm2-mm-manarea-depth).

@@

Timeseries clustering identified three clusters in reef relative cover timeseries for the whole GBR scale. A ‘low’ cluster of reefs with little recovery following initial declines, a medium cluster of reefs with median relative cover below 0.5 post 2028, and a high cluster of reefs with median relative cover above 0.5 until 2050 (@sfig-access-esm1-5-gbr-timeseries; @sfig-access-cm2-gbr-timeseries; @sfig-ec-earth3-veg-gbr-timeseries; @sfig-gfdl-cm4-gbr-timeseries; @sfig-noresm2-mm-gbr-timeseries).

The clusters identified with timeseries clustering at the GBR scale show similar patterns to clusters identified at bioregion and management area scales. Reef depth (@sfig-access-esm1-5-gbr-depth; @sfig-access-cm2-gbr-depth; @sfig-ec-earth3-veg-gbr-depth; @sfig-gfdl-cm4-gbr-depth; @sfig-noresm2-mm-gbr-depth) and total cover – DHW correlation (@sfig-access-esm1-5-gbr-dhwcor; @sfig-access-cm2-gbr-dhwcor; @sfig-ec-earth3-veg-gbr-dhwcor; @sfig-gfdl-cm4-gbr-dhwcor; @sfig-noresm2-mm-gbr-dhwcor) decrease, while the proportion of coral-habitable area initially occupied by corals increases (@sfig-access-esm1-5-gbr-initialproportion; @sfig-access-cm2-gbr-initialproportion; @sfig-ec-earth3-veg-gbr-initialproportion; @sfig-gfdl-cm4-gbr-initialproportion; @sfig-noresm2-mm-gbr-initialproportion), as the clusters decrease from ‘high’ to ‘low’ cover levels. There are no substantial differences between the clusters in mean DHW (@sfig-access-esm1-5-gbr-dhwmean; @sfig-access-cm2-gbr-dhwmean; @sfig-ec-earth3-veg-gbr-dhwmean; @sfig-gfdl-cm4-gbr-dhwmean; @sfig-noresm2-mm-gbr-dhwmean) and DHW tolerance (@sfig-access-esm1-5-gbr-dhwtol; @sfig-access-cm2-gbr-dhwtol; @sfig-ec-earth3-veg-gbr-dhwtol; @sfig-gfdl-cm4-gbr-dhwtol; @sfig-noresm2-mm-gbr-dhwtol) levels of reefs. Similarly, there are no substantial differences between clusters in connectivity metrics source – to – sink ratio and total connectivity strength (@sfig-access-esm1-5-gbr-logsosi; @sfig-access-cm2-gbr-logsosi; @sfig-ec-earth3-veg-gbr-logsosi; @sfig-gfdl-cm4-gbr-logsosi; @sfig-noresm2-mm-gbr-logsosi; @sfig-access-esm1-5-gbr-totalconnectivity; @sfig-access-cm2-gbr-totalconnectivity; @sfig-ec-earth3-veg-gbr-totalconnectivity; @sfig-gfdl-cm4-gbr-totalconnectivity; @sfig-noresm2-mm-gbr-totalconnectivity).

@@

Timeseries analysis has been used to identify three groups of reefs with similar declining (low), recovering (high), and intermediate (medium) coral cover trajectories throughout future scenarios. Under all GCMs most reefs suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. The differences between low, medium and high reef clusters in their timeseries are most notable under these two GCMs. At all analysis scales reef clusters show similar patterns in their reef characteristics such as depth and proportion of habitable area initially occupied.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced, and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to calculate the degree of protection provided to a reef based on the median depth. Our results suggest that within shallow reef ecosystems (depth < 30m), depth contributes greatly to reef resilience and the ability of the reef to maintain cover under heat stress. While similar findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses have often focussed on the potential of mesophotic reefs (> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific. 
<!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->

@@

Current reef restoration and intervention-investigation efforts on the Great Barrier Reef, such as deployment of coral seeding and fragment devices target reefs with suitable restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One of the main depth constraints is being able to visually identify the target site from the surface for deployment and ongoing monitoring, in combination with factors such as deployment method and tourism return value. Our results suggest that of the reefs analysed within the 1-10m depth range (1800 reefs) 74% (1328) of reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region scale), with many suffering large future declines in reef health under climate change and possibly limiting the potential of interventions to deliver recovery in these locations. Reef intervention planning is a complex and challenging decision making process [@anthony2020], and our results suggest that deployment site depth is an important factor to consider when assessing the long-term resilience of reef sites.

@@

Furthermore, our results suggest that the capacity of reefs to maintain healthy coral habitat by being above positive net carbonate budget levels is greater for reefs at deeper depths than shallow depths. While the maintanence of positive carbonate budget levels does not appear to be influenced by reef connectivity metrics. The maintanence of positive carbonate budgets is important for reef resilience under climate change.

@@

Further analyses could build on the findings of this study by analysing the impacts of prioritising reef depth in reef restoration intervention decision making. The ADRIA platform is developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and ecosystem scenario modelling into reef restoration decision analysis. The platform allows for the investigation of different reef characteristics to be targetted in decision making and models the evolving outcomes as a result of reef interventions at various locations. The capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of waves, cyclones, CoTS and currents into reef resilience investigations [@lin2023; @mcclanahan2005; @grimaldi2022; @liu2024]. Additionally, the investigation could be enhanced by incorporating future ocean acidification into considerations of coral carbonate budgets required for healthy habitat production.

@@

![Reef Degree Heating Week timeseries clustered within each management area, for each GCM analysed. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Each management area is represented five times (for each GCM), with each GCM displayed horizontally.](/../outputs/ADRIA_results/HighResCoralStress/figs/GCM_timeseries_plot.png){#fig-gcm-trajectories width=`{julia} timeseries_width`}

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to 2099. The carbonate budget thresholds assessed range from 10 to 20% of total reef area. Point colour represents log scaled reef total connectivity strength. Displays results from median timeseries using ACCESS-ESM1-5.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/log_total_strength_carbonate_scatter.png){#sfig-access-esm1-5-conn-carbonate width=`{julia} carb_width`}
